On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-05-16 Manuscript/manuscript.tex
cleaning of resolved comments

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the Men's 110 meter hurdle final of the 2022
World Track and Field Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun %scrutinized the results %and performed 
%non-statistical analysis of the data. They found 
concluded that the reaction time data from the 2022 World Championships 
seemed to be generally faster compared to the other datasets they looked at, but 
they did not perform any formal %meaningful 
statistical analyses. This paper questions the reaction time 
disqualification barrier, which is currently 0.1 seconds, to determine whether 
this is a reasonable threshold. We employ a generalized linear mixed model 
(GLMM) with a random %effect term which will be referred to as the 
venue effect in order to %try and 
model the data. Additionally, we employ a signed-rank test for clustered data to 
compare reaction times for the same athletes at different competitions. This 
matter needs to be addressed because %Devon Allen's 
disqualification based on allowable reaction time 
%could be a 
%problem that could 
will continue to be an issue in future world championships.

@@

\eds{also removed 2022 World Track and Field Championships from keywords and 
changed location of World in the title. World Track and Field Championships
has more Google search results than Track and Field World Championships, so
I opted to go with this. Also used plural Championships. Feel free to switch 
order or use singular if one or both of these is 
preferred for some reason.  }

@@

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championships 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year. 
\citep{johnson2022was}.
%came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
%reaction times came during the same championships for the men's 100 meter,
%men's 110 meter hurdle, and women's 100 meter for 13 competitions held between 2003 and 2022 
%\of{Cleaned up but may need additional work}
\of{Cleaned this up a little but still a little messy.}
\eds{is this better? (and still correct?)}
\of{Yes, I think it looks good}

%3)
%The International Association of Athletics Federations (also referred to
%as IAAF or World Athletics), which is the governing board for international running
%competitions, commissioned a study in 2009 regarding the reaction time 
%barrier and whether the threshold should be lowered. 
%The paper analyzed sprinters reaction times and the biological processes 
%that went into reacting to the starting gun and concluded 
%that athletes can react in as low as 0.08 seconds.  
%One study by Pain looked at
Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}. This 
contradicts the results of \citet{zhang2021correlation}, which indicated that 
year of competition has a significant effect on reaction times.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.001 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 World Track and Field Championships 
and at his physical peak.

@@

Despite their study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is confusing
as to why IAAF would commission a study only to disregard the results and not
make the necessary changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.
%Reaction time barrier has been examined before: cite reviews
%Then go into specifics and talk about people who wanted to change barrier
%General reviews are important
%People have studied the biological processes, cite a few references
%How barrier is picked by regulator
%Read abstract and into and summarize into about one sentence what each paper contributes
%1 and 2 can be merged but 3 need to stand alone

%4)
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this two different ways.  First we model the data, which is
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
%\eds{are you using covariates?  Data section
%implies that you are, but Results section does not}
%\of{Cleaned up confusion about covariates, but no none were used in the paper} 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 data differed significantly from other 
competitions.
Our investigation into reaction times is important not only because if its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
international and 2019 international vs 2022 international.  Should I instead say
that there are three types of data across 2 sections or something like that?}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix.%~\ref{sec:Appendix}.   %\eds{see earlier comment - 
%are you using any of these as covariates?}
%\of{removed covariate confusion, talk about womens a couple sentences down} 
Although women's data is available, we chose not to include women's data in this
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}.
%The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
%semifinal heat (of which there are either two
%or three each year) denoted by ``S", or final heat denoted by ``F".
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats %were not studied because athletes react faster as the 
%competition progresses 
\citep[e.g.,][]{zhang2021correlation}. 
%The variable 
%``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  
%The variable ``Batch" takes a numerical value and every heat of every round from 
%1999 to 2022 was assigned it's
%own number.  This was added because of suspected issues that reaction time may be
%correlated with other runners reacting quickly (one runner reacting quickly may
%result in other runners reacting quickly). There is no statistical meaning to
%how the batches were labeled, they were ordered chronologically by meet but
%reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
%111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
 % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
%\eds{if you are really not using ``H'' data at all, it does not appear necessary
%to spend time defining earlier.  You could just say directly when you define 
%stage that you only consider semifinals and finals and cite the Zhang paper for 
%rationale}
%\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
%use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few final heat observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.

The data under consideration is best visualized using side-by-side box plots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the reaction times in the 
2022 box plot for both 
the men's finals and men's pooled semifinal and final data are lower (indicating 
faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found median reaction times of 0.156 seconds 
for men when looking at data from 1999 to 2014.

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
Something that is extremely important to note in the data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Championships) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
This was desirable for World Athletics as false starts can make already long 
track meets more tedious, both for athletes and viewers watching the television 
broadcast.  There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championships there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.
  %\eds{I am not sure the previous sentence follows from the one preceding it. Why
% does World Athletics find faster reaction times desirable?  Do you mean that
%they find fewer false starts desirable?}

%\eds{Two comments about the previous sentence: 1. 
%it does not logically follow the previous sentence as there is no mention of 
%gender, and 2. if you are not using women's data, don't spend time talking about 
%it at the beginning of the data section.  Similar to my comment about 
%preliminary heats, you can move this sentence to the beginning of the Data 
%section to provide a rationale for why you do not include data for women.} 
%\of{Updates: I moved any talk about women's data to start of Data section, I think
%it flows much better now}

\subsection{Beyond World Championships}\label{sec:databeyond}
The motivation for looking beyond World Championships data came from an 
exploratory analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
%\eds{be consistent with how you refer to the US championship:
%``United States Track and Field Championships" and ``United States Track and Field 
%Championships" are both being used; please pick one and use that throughout}
%\of{Resolved; changed all USA to United States.}
From June 23-26 2022, the United States held its Track
and Field Championships at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championships a month later.  
%Thus, a baseline is able to be established for the four United States athletes who 
Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championships races compared to the United States Track and Field 
Championships races. Devon Allen raced three times at the United States Track and
Field Championships and three times at the World Track and Field Championships. 
His reaction times were substantially faster in the World Track and
Field Championships compared to the United States Track and Field Championships 
as were his
teammates. 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. 
%``United States" denotes
%the United States Track and Field Championships, ``World" denotes the World Track and Field
%Championships. All numbers listed in the table are the reaction
%time for the respective athletes in seconds.

@@

\of{Im not sure how to fix this but the spacing between the lines of the
table is different than the rest of the paper}
\eds{LaTeX keeps captions singlespaced automatically with the doublespace 
command; don't worry about it!}

>>>>

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the men's 110 meter hurdle final of the 2022
World Track and Field Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun %scrutinized the results %and performed 
%non-statistical analysis of the data. They found 
concluded that the reaction time data from the 2022 World Championships 
seemed to be generally faster compared to the other datasets they considered, 
but they did not perform any formal %meaningful 
statistical analyses. This paper questions the reaction time 
disqualification barrier, which is currently 0.1 seconds, to determine whether 
this is a reasonable threshold. We employ a generalized linear mixed model 
(GLMM) with a random %effect term which will be referred to as the 
venue effect in order to %try and 
model the reaction time data. Additionally, we employ a signed-rank test for 
clustered data to 
compare reaction times for the same athletes at different competitions. This 
matter needs to be addressed because %Devon Allen's 
disqualification based on allowable reaction time 
%could be a 
%problem that could 
will continue to be an issue in future world championships.

@@

%\eds{also removed 2022 World Track and Field Championships from keywords and 
%changed location of World in the title. World Track and Field Championships
%has more Google search results than Track and Field World Championships, so
%I opted to go with this. Also used plural Championships. Feel free to switch 
%order or use singular if one or both of these is 
%preferred for some reason.  }

@@

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championships 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.
%came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
%reaction times came during the same championships for the men's 100 meter,
%men's 110 meter hurdle, and women's 100 meter for 13 competitions held between 2003 and 2022 
%\of{Cleaned up but may need additional work}
%\of{Cleaned this up a little but still a little messy.}
%\eds{is this better? (and still correct?)}
%\of{Yes, I think it looks good}

%3)
%The International Association of Athletics Federations (also referred to
%as IAAF or World Athletics), which is the governing board for international running
%competitions, commissioned a study in 2009 regarding the reaction time 
%barrier and whether the threshold should be lowered. 
%The paper analyzed sprinters reaction times and the biological processes 
%that went into reacting to the starting gun and concluded 
%that athletes can react in as low as 0.08 seconds.  
%One study by Pain looked at
Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
\eds{are hurdlers generally considered a subset of sprinters or are they 
considered a separate type of athlete?  Specifically did the study above 
recommend the barrier be lowered for both sprinters and hurdlers?
You also refer to sprinters elsewhere in the manuscript.  Are those instances
meant to refer to both sprinters and hurdlers, as well?}
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}. This 
contradicts the results of \citet{zhang2021correlation}, which indicated that 
year of competition has a significant effect on reaction times.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.001 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 World Track and Field Championships 
and likely at his physical peak.

@@

Despite their study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is unclear
why IAAF would commission a study only to disregard the results and not
make the recommended changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.
%Reaction time barrier has been examined before: cite reviews
%Then go into specifics and talk about people who wanted to change barrier
%General reviews are important
%People have studied the biological processes, cite a few references
%How barrier is picked by regulator
%Read abstract and into and summarize into about one sentence what each paper contributes
%1 and 2 can be merged but 3 need to stand alone

%4)
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 reaction times differed significantly from 
those of other competitions.
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
international and 2019 international vs 2022 international.  Should I instead say
that there are three types of data across 2 sections or something like that?}
\eds{I think this is fine here, but it is not at all clear that there are two
subsections in the rank-based results for 2022 national vs 2022
international and 2019 international vs 2022 international.}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix. %~\ref{sec:Appendix}.   %\eds{see earlier comment - 
%are you using any of these as covariates?}
%\of{removed covariate confusion, talk about womens a couple sentences down} 
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
%The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
%semifinal heat (of which there are either two
%or three each year) denoted by ``S", or final heat denoted by ``F".
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats %were not studied because athletes react faster as the 
%competition progresses 
\citep[e.g.,][]{zhang2021correlation}. 
%The variable 
%``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  
%The variable ``Batch" takes a numerical value and every heat of every round from 
%1999 to 2022 was assigned it's
%own number.  This was added because of suspected issues that reaction time may be
%correlated with other runners reacting quickly (one runner reacting quickly may
%result in other runners reacting quickly). There is no statistical meaning to
%how the batches were labeled, they were ordered chronologically by meet but
%reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
%111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
 % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
%\eds{if you are really not using ``H'' data at all, it does not appear necessary
%to spend time defining earlier.  You could just say directly when you define 
%stage that you only consider semifinals and finals and cite the Zhang paper for 
%rationale}
%\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
%use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few final heat observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.

The data under consideration is best visualized using side-by-side boxplots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the reaction times in the 
2022 boxplot for both 
the men's finals and men's pooled semifinal and final data are lower (indicating 
faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found median reaction times of 0.156 seconds 
for men when looking at data from 1999 to 2014.

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
\eds{With the following newly added sentences, I think we can leave this paragraph here.}
We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period. 
%Something that is extremely important to note in the data is the inconsistency 
%of World Athletics in terms of how they penalize false starts.  
From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Championships) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
%This was desirable for World Athletics as false starts can make already long 
%track meets more tedious, both for athletes and viewers watching the television 
%broadcast.  
%\eds{I think we lost something here with all of the edits so I removed the 
%previous sentence. What was desirable for World Athletics -- the 2007-2009 rule
% change?  Why was this desirable?  The rule change would seem to make the event
% longer and not shorter? } 
There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female false starts at the 2009 World 
Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championships there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.

\subsection{Beyond World Championships}\label{sec:databeyond}
The motivation for looking beyond World Championships data came from an 
exploratory analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
%\eds{be consistent with how you refer to the US championship:
%``United States Track and Field Championships" and ``United States Track and Field 
%Championships" are both being used; please pick one and use that throughout}
%\of{Resolved; changed all USA to United States.}
From June 23-26 2022, the United States held its Track
and Field Championships at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championships a month later.  
%Thus, a baseline is able to be established for the four United States athletes who 
Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championships races compared to the United States Track and Field 
Championships races. Devon Allen raced three times at the United States Track and
Field Championships and three times at the World Track and Field Championships. 
His reaction times were substantially faster in the World Track and
Field Championships compared to the United States Track and Field Championships, 
as were his
teammates. 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts.

@@

%\of{Im not sure how to fix this but the spacing between the lines of the
%table is different than the rest of the paper}
%\eds{LaTeX keeps captions singlespaced automatically with the doublespace 
%command; don't worry about it!}

This dataset is too small to perform any meaningful statistical analysis, but
the data were expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  \eds{this is where you can put the relevant 
information about the other competitions that was commented out in the 
introduction.}
\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\of{Is the one below what you're talking about?}
\eds{Yes, this is the part from the intro I was referring to, but you need to 
edit this paragraph now so it all works together.}
First we can compare the reaction times of athletes who
competed at the World Track and Field Championships in both 2019 and 2022, and 
second we can compare how athletes performed at the 2022 championships relative 
to their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.
